---
title: "QC and Filtering"
author: "ItunuBabatunmise"
date: "2026-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)
library(Matrix)
library(scales)
library(cowplot)
library(RCurl)
```

```{r}
# Create output directories if they don't exist
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)
```

```{r}
# Read in all the samples
# Get all folders inside the data directory
data_dirs <- list.dirs("Data", recursive = FALSE)

# Create an empty list to store Seurat objects
seurat_list <- list()

# Loop through each folder
for (dir_path in data_dirs) {
  
  sample_name <- basename(dir_path)
  
  # Read the matrix
  mtx_obj <- ReadMtx(
    mtx = file.path(dir_path, "matrix.mtx.gz"),
    features = file.path(dir_path, "features.tsv.gz"),
    cells = file.path(dir_path, "barcodes.tsv.gz")
  )
  
  # Create Seurat object
  seurat_obj <- CreateSeuratObject(counts = mtx_obj, project = sample_name)
  
  # Store using folder name as list key
  seurat_list[[sample_name]] <- seurat_obj
  
  cat("Processed:", sample_name, "\n")
}
```

```{r}
# Check metadata for each sample
head(seurat_list[["S1"]]@meta.data)
head(seurat_list[["S2"]]@meta.data)
head(seurat_list[["S3"]]@meta.data)
head(seurat_list[["S4"]]@meta.data)
head(seurat_list[["S5"]]@meta.data)
```

```{r}
# Create a merged Seurat object to combine all samples
merged_seurat <- merge(
  x = seurat_list[[1]],
  y = seurat_list[-1],
  add.cell.ids = c("S1", "S2", "S3", "S4", "S5")
)
```

```{r}
# Check that the merged object has the appropriate sample-specific prefixes
head(merged_seurat@meta.data)
tail(merged_seurat@meta.data)
```

# Metadata fields:
# orig.ident: sample identity (defaults to project name)
# nCount_RNA: number of UMIs per cell
# nFeature_RNA: number of genes detected per cell

```{r}
# Explore merged metadata
View(merged_seurat@meta.data)
```

# Generating Quality Metrics
```{r}
# Compute percent mito ratio
merged_seurat$mitoRatio <- PercentageFeatureSet(object = merged_seurat, pattern = "^MT-")
merged_seurat$mitoRatio <- merged_seurat@meta.data$mitoRatio / 100
```

```{r}
# Create metadata dataframe
metadata <- merged_seurat@meta.data

# Add cell IDs to metadata
metadata$cells <- rownames(metadata)

# Rename columns
metadata <- metadata %>%
  dplyr::rename(
    seq_folder = orig.ident,
    nUMI = nCount_RNA,
    nGene = nFeature_RNA
  )
```
# Additional Metrics
# Number of genes detected per UMI. Add number of genes per UMI for each cell to metadata
```{r}
# This metric gives an idea of the complexity of the dataset
# (more genes detected per UMI = more complex data)
metadata$log10GenesPerUMI <- log10(metadata$nGene) / log10(metadata$nUMI)
```

```{r}
# Create sample column- to reduce the long sample name to S1...
metadata$sample <- NA
metadata$sample[which(str_detect(metadata$cells, "^S1_"))] <- "S1"
metadata$sample[which(str_detect(metadata$cells, "^S2_"))] <- "S2"
metadata$sample[which(str_detect(metadata$cells, "^S3_"))] <- "S3"
metadata$sample[which(str_detect(metadata$cells, "^S4_"))] <- "S4"
metadata$sample[which(str_detect(metadata$cells, "^S5_"))] <- "S5"
```

```{r}
# Add metadata back to Seurat object
merged_seurat@meta.data <- metadata

# Save merged Seurat object
save(merged_seurat, file = "Data/merged_filtered_seurat.RData")
```

# Cell Counts
Unique cellular barcodes detected per sample.

```{r}
p <- metadata %>%
  ggplot(aes(x = sample, fill = sample)) +
  geom_bar() +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Number of Cells")

ggsave("results/figures/number_of_cells.pdf", plot = p, width = 8, height = 6)
p
```

# UMI Counts (Transcripts) per Cell

```{r}
p <- metadata %>%
  ggplot(aes(color = sample, x = nUMI, fill = sample)) +
  geom_density(alpha = 0.2) +
  scale_x_log10() +
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500) +
  ggtitle("Distribution of UMIs per Cell")

ggsave("results/figures/UMI_density_per_cell.pdf", plot = p, width = 8, height = 6)
p
```

# Genes Detected per Cell

```{r}
p1 <- metadata %>%
  ggplot(aes(color = sample, x = nGene, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  geom_vline(xintercept = 4000) +
  ylab("Cell density") +
  ggtitle("Distribution of Genes Detected per Cell")

ggsave("results/figures/gene_density_per_cell.pdf", plot = p1, width = 8, height = 6)
p1
```

```{r}
p2 <- metadata %>%
  ggplot(aes(x = sample, y = log10(nGene), fill = sample)) +
  geom_boxplot() +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("NCells vs NGenes")

ggsave("results/figures/nCells_vs_nGenes_boxplot.pdf", plot = p2, width = 8, height = 6)
p2
```

# UMIs vs. Genes Detected

```{r}
p <- metadata %>%
  ggplot(aes(x = nUMI, y = nGene)) +
  geom_point(alpha = 0.5) +
  stat_smooth(method = lm, color = "red") +
  scale_x_log10() +
  scale_y_log10() +
  theme_classic() +
  geom_vline(xintercept = 500) +
  geom_hline(yintercept = 250) +
  facet_wrap(~sample) +
  ggtitle("Correlation Between UMIs and Genes Detected")

ggsave("results/figures/UMI_vs_Genes_correlation.pdf", plot = p, width = 10, height = 8)
p
```
```{r}
# Mitochondrial Counts Ratio
p <- metadata %>%
  ggplot(aes(color = sample, x = mitoRatio, fill = sample)) +
  geom_density(alpha = 0.2) +
  scale_x_log10() +
  theme_classic() +
  geom_vline(xintercept = 0.2) +       
  ylab("Cell density") +
  ggtitle("Mitochondrial Counts Ratio")  
ggsave("results/figures/Mitochondrial_counts_ratio.pdf", plot = p, width = 10, height = 8)
p
```
# Complexity
visualizing the genes detected per UMI
```{r}
p <- metadata %>%
  ggplot(aes(x = log10GenesPerUMI, color = sample, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8) +
  ylab("Cell density") +
  ggtitle("Gene Expression Complexity per Cell")

ggsave("results/figures/genes_per_UMI_complexity.pdf", plot = p, width = 8, height = 6)
p
```

# Gene Level Filtering
Removing genes with zero expression across all cells and genes expressed
in fewer than 10 cells, as these reduce average expression without adding
meaningful information.

```{r}
# View the layers in your data
Layers(merged_seurat)
```
```{r}
# Join layers before filtering
merged_seurat <- JoinLayers(merged_seurat)

# Extract counts
counts <- GetAssayData(object = merged_seurat, layer = "counts")

# Keep genes expressed in 10 or more cells
keep_genes <- Matrix::rowSums(counts > 0) >= 10

# Subset Seurat object
merged_seurat <- subset(
  merged_seurat,
  features = rownames(counts)[keep_genes]
)
```

```{r}
# Save filtered Seurat object
save(merged_seurat, file = "Data/seurat_filteredQC.RData")
```

# Re-assess QC Metrics After Filtering

```{r}
metadata_clean <- merged_seurat@meta.data
```

```{r}
# Visualize the number of cell counts per sample
p <- metadata_clean %>%
  ggplot(aes(x = sample, fill = sample)) +
  geom_bar() +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("NCells After Filtering")

ggsave("results/figures/post_filter_ncells.pdf", plot = p, width = 8, height = 6)
p
```

```{r}
# Visualize the number UMIs/transcripts per cell
p <- metadata_clean %>%
  ggplot(aes(color = sample, x = nUMI, fill = sample)) +
  geom_density(alpha = 0.2) +
  scale_x_log10() +
  theme_classic() +
  ylab("Cell density") +
  geom_vline(xintercept = 500) +
  ggtitle("UMI Density After Filtering")

ggsave("results/figures/post_filter_UMI_density.pdf", plot = p, width = 8, height = 6)
p
```

```{r}
# Visualize the distribution of genes detected per cell via histogram
p1 <- metadata_clean %>%
  ggplot(aes(color = sample, x = nGene, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  scale_x_log10() +
  geom_vline(xintercept = 300) +
  ggtitle("Gene Density After Filtering")

ggsave("results/figures/post_filter_gene_density.pdf", plot = p1, width = 8, height = 6)
p1
```

```{r}
#Visualize the distribution of genes detected per cell via boxplot
p2 <- metadata_clean %>%
  ggplot(aes(x = sample, y = log10(nGene), fill = sample)) +
  geom_boxplot() +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("NCells vs NGenes After Filtering")

ggsave("results/figures/post_filter_nCells_vs_nGenes.pdf", plot = p2, width = 8, height = 6)
p2
```

```{r}
# Visualize the overall complexity
p <- metadata_clean %>%
  ggplot(aes(x = log10GenesPerUMI, color = sample, fill = sample)) +
  geom_density(alpha = 0.2) +
  theme_classic() +
  geom_vline(xintercept = 0.8) +
  ggtitle("Complexity After Filtering")

ggsave("results/figures/post_filter_complexity.pdf", plot = p, width = 8, height = 6)
p
```
